\documentclass[11pt]{article}

\input{../shared/preamble.tex}

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{cleveref}

\title{Winning Probabilities as Credit:\\
       Fast, Redundancy-Aware Attribution}
\author{Peter Cotton}
\date{\today}

\newcommand{\E}{\mathbb{E}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\DeclareMathOperator*{\argmaxop}{arg\,max}

\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{definition}{Definition}

\begin{document}
\maketitle

\begin{abstract}
The probability that a competitor wins a noisy race---the Thurstonian choice
probability, equivalently the gradient of an expected-maximum potential---is a
rule for attributing credit among correlated contributors. It is a probability share
($\sum_i w_i = 1$), symmetric, and \emph{redundancy-aware} (contributors that move
together share a single contributor's credit rather than double-counting); it is
differentiable in the contributors' abilities, it updates online, and it costs
$O(Mn)$ in a single Monte-Carlo pass with no coalition enumeration. We place it next
to the Shapley value while keeping the units straight: the winning probability is the
\emph{gradient field} of the expected-maximum potential, and its diagonal line
integral (weighted by the abilities) is an Aumann--Shapley value of the
ability-scaling game, but the probability itself is not the Shapley value of the
membership game---a cousin, not a drop-in, and in different units. The practical separation is one of \emph{form}: the race returns
the gradient credit in a single differentiable pass, where a generic Shapley
evaluator faces $2^n$ coalitions (the specialized max-game value is cheaper, but not
a one-pass gradient). We show that the headline redundancy property---an even split
of credit among near-duplicates---is a property of the \emph{calibrated} equal-ability
race rather than of raw scores, and we demonstrate the rule on forecast combination
and feature attribution. This develops the credit reading of the construction in the
companion paper on Thurstone portfolios.
\end{abstract}

\section{Introduction}
\label{sec:intro}

Attributing credit for a collective outcome among contributors that overlap is an
old problem with a canonical answer, the Shapley value \parencite{shapley1953}, and
a canonical difficulty, its cost: the exact value sums marginal contributions over
all $2^n$ coalitions, and the practical estimators sample permutations
\parencite{strumbelj2014} or, in machine learning, fit a local surrogate (SHAP,
\textcite{lundberg2017}). Correlated contributors compound the difficulty: a
contributor and a near-duplicate should \emph{share} credit, not each receive it in
full, and many cheap importance measures (marginal screening, regression
coefficients, permutation importance) either double-count the pair or split it
arbitrarily.

We study a different credit rule with a single, cheap evaluation. Give each
contributor $i$ a latent ability $\theta_i$ and a noisy performance
$X_i = \theta_i + \eta_i$ under a joint law $S$; the credit of $i$ is the
probability that it wins,
\begin{equation}
  w_i \;=\; \Prob\!\Big( X_i = \max_j X_j \Big).
  \label{eq:winprob}
\end{equation}
We follow the lowest-\emph{score}-wins convention of the \texttt{thurstone} package
and \textcite{cotton2021}: $\theta_i$ is a \emph{strength}, the negated score, so the
lowest-score contributor is the $\argmaxop$ winner in \eqref{eq:winprob}; in the
loss-based applications of \Cref{sec:apps} the score is a loss and the lowest-loss
contributor wins. This is the Thurstonian choice probability, and $w = W_S(\theta)$
is the gradient of an expected-maximum potential (\Cref{sec:rule}). Read as credit it has
three things going for it at once. It is redundancy-aware by construction: two
contributors that move together compete with \emph{each other}, so a co-moving pair
collapses toward a single winner and shares one contributor's credit. It is
differentiable in $\theta$, so it composes with gradient-based learning. And it
costs one Monte-Carlo pass, $O(Mn)$, returning all $n$ shares at once with no
coalition enumeration.

Our contributions are (i) the precise relationship to the Shapley value: the
winning probability is the gradient of the same expected-maximum potential whose
\emph{diagonal integral} is an Aumann--Shapley value
\parencite{owen1972, aumannshapley1974}, but it is not itself the Shapley value of
the membership game (\Cref{sec:shapley}); (ii) the computational separation,
$O(Mn)$ versus $O(2^n)$, quantified (\Cref{sec:complexity}); and (iii) the
clarification that the headline redundancy property---an \emph{even} split among
near-duplicates---belongs to the \emph{calibrated} equal-ability race, not to raw
contribution scores (\Cref{sec:redundancy}). We close with two applications,
forecast combination and feature attribution (\Cref{sec:apps}).

\section{The credit rule}
\label{sec:rule}

Let $S$ be a centered joint law on $\R^n$ (the \emph{simulation}) and define the
expected-maximum potential and its gradient,
\begin{equation}
  G_S(\theta) \;=\; \E_{\eta\sim S}\Big[\max_i (\theta_i + \eta_i)\Big],
  \qquad
  W_S(\theta) \;=\; \nabla G_S(\theta) \in \Delta,
  \label{eq:potential}
\end{equation}
where $\Delta$ is the simplex. $G_S$ is convex for any $S$; when $S$ is continuous and
the usual dominated-differentiation conditions hold, the maximizer is a.s.\ unique,
$G_S$ is differentiable, and $\nabla_i G_S(\theta) = \Prob(i = \argmaxop_j X_j) = w_i$,
recovering \eqref{eq:winprob} (the Williams--Daly--Zachary / perturbed-maximum
identity \parencite{williams1977, mcfadden1981, berthet2020}); for Gaussian $S$,
$G_S$ is $C^\infty$. Because $G_S(\theta + c\mathbf 1) = G_S(\theta) + c$, gradients
and Jacobians live on the tangent space $T\Delta = \{v : \mathbf 1^\top v = 0\}$.

To use \eqref{eq:winprob} as an attribution we must fix the abilities. Given a
\emph{benchmark} $w^0 \in \operatorname{int}\Delta$ of marginal credits and a
reference law $S_0$, calibration inverts the gradient map,
\begin{equation}
  \theta^0 \;=\; W_{S_0}^{-1}(w^0) \quad (\text{mod } \mathbf 1),
  \label{eq:calib}
\end{equation}
which is well posed: for Gaussian $S_0 = \mathcal N(0, C_0)$ with $C_0 \succ 0$,
$G_{C_0}$ is a Legendre-type convex function on the quotient
$\R^n/\!\operatorname{span}\{\mathbf 1\}$---smooth, strictly convex, and steep toward
the boundary of its image---so $W_{C_0} = \nabla G_{C_0}$ is a smooth bijection onto
$\operatorname{int}\Delta$; the diagonal case admits a linear-time, Monte-Carlo-free
inverse (the Fast Ability Transform, \textcite{cotton2021}). A zero target weight
sends its ability to $-\infty$, so such contributors are dropped or floored. The
credit rule is then
\begin{equation}
  T_{S_0\to S_1}(w^0) \;=\; W_{S_1}\!\big(W_{S_0}^{-1}(w^0)\big),
  \label{eq:tilt}
\end{equation}
the winning probabilities of the same abilities re-evaluated under a
dependence-rich law $S_1$. With $S_1 = S_0$ the benchmark is reproduced; with $S_1$
carrying the contributors' empirical dependence, \eqref{eq:tilt} re-weights the
benchmark for redundancy.

\section{Relationship to the Shapley value}
\label{sec:shapley}

The winning probability is naturally compared to the Shapley value, and the
relationship is real but easy to overstate, because the two are credit in different
\emph{units}. The winning probabilities are \emph{probability shares},
$\sum_i w_i = 1$. The Shapley value allocates a cooperative game's grand value: for
\begin{equation}
  v(\mathcal S) \;=\; \E\Big[\max_{i\in\mathcal S} X_i\Big],
  \qquad v(\varnothing) = 0,
  \label{eq:game}
\end{equation}
with grand value $v(N) = \E[\max_i X_i] = G_S(\theta)$, the Shapley shares sum to
$v(N)$---a payoff in the units of $X$, not to one. So $w$ is \emph{not} a Shapley
value of $v$, and the two obey different efficiency conditions, $\sum_i w_i = 1$
versus $\sum_i \phi_i = v(N)$. The honest connection is a gradient/integral one,
through the ability-scaling game.

\begin{proposition}[Gradient, integral, and the membership value]
\label{prop:bridge}
Let $S$ be continuous with the usual dominated-differentiation conditions, and define
the ability-scaling game $f(x) = G_S(x\odot\theta) - G_S(0)$ for $x\in[0,1]^n$, where
$x\odot\theta$ scales ability $i$ by $x_i$.
\begin{enumerate}
\item[\textup{(i)}] \emph{Gradient.} The winning-probability vector is the endpoint
  gradient, $w = \nabla G_S(\theta)$, with $\sum_i w_i = 1$.
\item[\textup{(ii)}] \emph{Aumann--Shapley.} The diagonal Aumann--Shapley value of
  $f$ \parencite{aumannshapley1974} is
  $\mathrm{AS}_i = \int_0^1 \partial_{x_i} f(t\mathbf 1)\,dt
   = \theta_i \int_0^1 W_S(t\theta)_i\,dt$, in payoff units; it is efficient,
  $\sum_i \mathrm{AS}_i = G_S(\theta) - G_S(0)$, and symmetric. The
  winning-probability field $\theta\mapsto W_S(\theta)$ is the \emph{gradient} whose
  diagonal line integral (weighted by $\theta$) is $\mathrm{AS}$; $w$ is the endpoint
  gradient, not the integral.
\item[\textup{(iii)}] \emph{Membership Shapley.} The Shapley value of \eqref{eq:game}
  is, by Owen's multilinear extension \parencite{owen1972}, the diagonal integral
  over a random coalition $T$ of the marginal $v(T\cup\{i\}) - v(T)$, which equals
  $\E[(X_i - \max_{j\in T} X_j)^+]$ for $T\neq\varnothing$ and $\E[X_i]$ for
  $T = \varnothing$. This payoff-valued, coalition-averaged quantity is a different
  object from the probability $w_i = \Prob(X_i = \max_N)$, and the two generally
  differ (they agree in the fully symmetric case).
\end{enumerate}
\end{proposition}
\begin{proof}
(i) is the Williams--Daly--Zachary / perturbed-maximum identity
\parencite{williams1977, berthet2020}, valid under the stated regularity. For (ii),
the chain rule and (i) give $\partial_{x_i} f(x) = \theta_i\, W_S(x\odot\theta)_i$;
on the diagonal this is the stated integrand, and efficiency is the fundamental
theorem of calculus,
$\sum_i \mathrm{AS}_i = \int_0^1 \tfrac{d}{dt} G_S(t\theta)\,dt = G_S(\theta)-G_S(0)$;
symmetry follows because exchanging equal-ability contributors leaves $G_S$
invariant. For (iii), when $T\neq\varnothing$,
$v(T\cup\{i\}) - v(T) = \E[\max(\max_{j\in T} X_j, X_i)] - \E[\max_{j\in T} X_j]
= \E[(X_i - \max_{j\in T} X_j)^+]$, while $v(\{i\}) - v(\varnothing) = \E[X_i]$;
Owen's identity integrates this over $T$. It is payoff-valued and averages over
coalitions, unlike the grand-coalition probability $w_i$.
\end{proof}

So the winning probability is the \emph{gradient field} of the expected-maximum
potential, and its diagonal line integral (weighted by the abilities) is an
Aumann--Shapley value of the ability-scaling game. It is not itself a Shapley value
of the membership game $v$---it is a probability share, in different units---and the
two agree only under full symmetry. What can be said axiomatically is about the
probability share directly, and it should not be dressed in the Shapley axioms (which
are conditions on a payoff value of $v$).

\begin{proposition}[Properties of the probability-share credit]
\label{prop:axioms}
The credit $w = \nabla G_S(\theta)$ satisfies \emph{probabilistic efficiency}
($\sum_i w_i = 1$); \emph{argmax symmetry} (interchangeable contributors receive
equal credit); and \emph{argmax nullity} (a contributor almost surely never the
maximizer receives $w_i = 0$ and leaves the others' credit unchanged). It is not a
linear functional of the game, so it is not additive in the Shapley sense.
\end{proposition}
\begin{proof}
Probabilistic efficiency and $w_i \ge 0$ are immediate. Argmax symmetry holds because
exchanging two contributors with the same marginal and dependence leaves $G_S$, hence
$\nabla G_S$, invariant. For argmax nullity, if $X_i \le \max_{j\neq i} X_j$ a.s.\
then $i$ is never the argmax, so $w_i = 0$, and $\max_N X = \max_{N\setminus\{i\}} X$
leaves the others unchanged. Non-additivity is direct: $w = \nabla G_S$ is the
gradient of a convex potential, nonlinear in $\theta$ (strictly convex only on the
quotient $\R^n/\!\operatorname{span}\{\mathbf 1\}$, being affine along $\mathbf 1$),
so the map from game to credit is not linear ($w(\theta+\theta') \neq w(\theta) +
w(\theta')$---the right-hand side need not even lie on the simplex).
\end{proof}

These are the right analogues: they are statements about a probability share, not the
Shapley efficiency or null-player axioms, which concern a payoff value of $v$.
Trading the linearity that pins down the Shapley value is what lets the credit be a
cheap, differentiable gradient.

\section{Computational complexity}
\label{sec:complexity}

The practical case for the winning probability as credit is its cost, stated against
the right baseline. For a \emph{generic} black-box cooperative game, exact Shapley
enumerates $2^n$ coalitions, and the usual recourse is permutation sampling,
$O(RMn)$ at $O(1/\sqrt R)$ error. But the expected-maximum game \eqref{eq:game} is
not black-box: on an empirical sample $X^{(1)},\dots,X^{(M)}$ the max-game Shapley
value has order-statistic structure and can be computed in about $O(Mn\log n)$
without enumerating coalitions. So the race's edge over a \emph{specialized} max-game
Shapley algorithm is not exponential. What the race buys is different: it returns the
endpoint-gradient credit---the probability share---in a single $\argmaxop$ pass over
the $M$ draws, $O(Mn)$, with no coalition or permutation machinery, and as the
gradient of a smooth potential it is differentiable in the abilities and supports
cheap online updates. Against a generic Shapley evaluator the gap is the exponential
one; against the best specialized one it is the constant factor of a single pass,
plus differentiability. The race is \emph{exact for the empirical simulation law}
defined by the $M$ draws; for the population law it carries the usual $O(M^{-1/2})$
Monte-Carlo error.

\section{Redundancy, and the calibration condition}
\label{sec:redundancy}

The reason to prefer this credit rule over a correlation-blind one is redundancy.

\begin{proposition}[Redundancy consistency at fixed abilities]
\label{prop:redundancy}
If contributors $i,j$ have equal ability and \emph{identical} performances
($X_i = X_j$ a.s.), then either wins exactly when the other does; with symmetric
tie-splitting each receives half of the single-contributor winning probability, and
the pair's combined credit equals that of one contributor.
\end{proposition}

This is a property of the race \emph{at fixed, equal abilities}. It does not follow
for an arbitrary vector of scores fed through $\argmaxop$: if near-duplicates carry
slightly unequal scores, the largest wins almost every draw and the split is
winner-take-all. The fix is calibration \eqref{eq:calib}: set the benchmark $w^0$ to
the contributors' \emph{marginal} credit---equal across genuine duplicates---then
tilt under their empirical dependence, so the pair is de-duplicated and split
evenly. The price is that the de-duplicated weight redistributes in part to weak
contributors; spreading credit and concentrating it cannot both be exact.

\section{Applications}
\label{sec:apps}

\paragraph{Forecast combination.}
Supplying the race with the base forecasters' per-period \emph{losses}
$L_{k,t}$ (so the lowest-loss forecaster wins) yields combination weights that are a
redundancy-aware credit split over models. On the M4 competition this is a promising
but inconclusive direction (the companion paper reports that study): in the
small-sample regime it is competitive with, and the most stable member of, the
combination family, but we claim no out-of-sample superiority. The performance
variable must be a loss, not a signed error, or the ``winner'' is merely the most
negatively biased forecaster.

\paragraph{Feature attribution.}
For model interpretability the contributors are features and the credit is a
redundancy-aware importance. Raw contribution scores give a quick screen but are
winner-take-all among correlated features; the calibrated race delivers the even
split that correlated-feature attribution wants, far more cheaply than a
coalition-based method. Unlike SHAP it returns a share of wins rather than an
additive decomposition of a single prediction---a global importance, not a local
explanation.

\paragraph{Variable selection.}
Thresholded, the same importance is a selection rule, and the natural competitor for
\emph{stable} selection under collinearity is stability selection
\parencite{meinshausen2010}, which ranks features by lasso selection frequency over
subsamples. On collinear designs the calibrated race is the more stable of the two:
across data resamples its per-duplicate credit varies several-fold less than both a
single lasso (which picks one copy and flips) and stability selection, while
splitting redundant features evenly---and it does so in one pass rather than a bag of
lasso fits. The costs are a softer signal/noise contrast, since some credit leaks to
weak features, and a higher per-call price than a single fit.

\section{Discussion}
\label{sec:discussion}

The winning probability is a fast, differentiable, online, redundancy-aware credit
rule---a cousin of the Shapley value with an exact Aumann--Shapley reading on the
ability game, and a large computational advantage where redundancy-aware credit is
what is wanted. The distinction from Shapley is best read as a choice of question,
not an approximation: Shapley measures \emph{coalitional fairness}---a contributor's
average marginal value as coalitions are assembled, a counterfactual-historical
quantity---while the winning probability measures \emph{field relevance}, how often a
contributor is uniquely the best in the field one actually has. For legal or moral
credit the coalitional average is the right object; for live selection---ensemble
weighting, feature screening, routing, ``which signal do I rely on now?''---the
full-field endpoint question is the operational one, and it is the one the race
answers. Its redundancy guarantee is exact for the calibrated equal-ability
race, which is the construction to use; raw scores give only a winner-take-all
screen. Its axiomatic standing is settled by \Cref{prop:axioms}: it keeps
efficiency, symmetry, and the null-player property, trading away only additivity.
As a variable-selection rule it is the stable, redundancy-aware option of
\Cref{sec:apps}. Its choice-sensitivity Jacobian $\nabla_\theta w$ turns out to be
well-conditioned even when the covariance is not: numerically its condition number
stays near $1$ while the covariance's grows into the hundreds, so the
ill-conditioning that defeats $\Sigma^{-1}$ never enters the choice-space inverse and
calibration needs no preconditioning. $J$ is also informative in its own right: it is
a \emph{substitutability matrix}---$J_{ij}$ measures how much contributors $i,j$
compete for the same winning event---so clustering it recovers the redundancy groups
\emph{endogenously}, from the allocation object itself, and in choice geometry rather
than raw correlation. Two contributors can be highly correlated yet not substitutes
in the race if one dominates and the other never wins; the correlation matrix cannot
tell these apart but $J$ can. We leave this interpretability layer to future work. The cost that does remain---the de-duplicated
credit that leaks to weak contributors, the one weakness recurring across the
attribution and selection experiments---is controlled by a concentration
(inverse-temperature) parameter $\beta$: scaling the calibrated strengths by $\beta$
drives a sub-maximal contributor's credit to zero as $\beta\to\infty$ while leaving
the even split among genuine duplicates intact, trading the leak against the
smoothness of the map (developed in the companion paper). A complementary device is
an \emph{outside option}: a null competitor whose winning probability $w_0$ absorbs
mass no real contributor deserves (an interpretable baseline---unexplained credit, or
cash / market-mode in the portfolio reading). Empirically the null preserves the even
split and softens the leak with a mild bias toward weak contributors, but it shaves
mass roughly in proportion to share, so it is a smooth interpretable floor rather than
a targeted leak remover; temperature is the sharper tool. The whole construction is,
in one line, a replacement of additive marginal contribution by the probability of
being the \emph{non-redundant winner in the actual field}. It is implemented in the
open-source \texttt{allocation} package, built on \texttt{thurstone}.

\printbibliography

\end{document}
